---
title: "Supervision 3"
format: html
---

## Lab 1 — Data Prep for Variable Selection (Hitters)

### 🎯 Learning outcomes
- Load the **Hitters** data and inspect dimensions.
- Handle **missing values** (Salary) to avoid errors in model fitting.
- Get comfortable with the packages we'll use: **ISLR** and **leaps**.

### 🧰 Prerequisites
- R and RStudio installed.
- Packages: `ISLR`, `leaps`.

```{r}
# install.packages(c("ISLR","leaps")) # if needed
library(ISLR)
data(Hitters)

# Inspect
names(Hitters)
dim(Hitters)          # rows x columns
sum(is.na(Hitters$Salary))

# Drop rows with any NA (Salary has some NAs)
Hitters <- na.omit(Hitters)
dim(Hitters)
sum(is.na(Hitters))
```
::: callout-note
We remove rows with missing **Salary** because selection routines (like `regsubsets`) require complete cases.
:::

---

## Lab 2 — Best Subset Selection

### 🎯 Learning outcomes
- Fit **best subset** models with `regsubsets()`.
- Retrieve the **best model** for each size.
- Inspect classical metrics: **RSS**, **R²**, **Adjusted R²**, **Cp**, **BIC**.

```{r}
library(leaps)

# Exhaustive search over all subsets
regfit.full <- regsubsets(Salary ~ ., data = Hitters, nvmax = 19)
reg.summary <- summary(regfit.full)

# What does summary() return?
names(reg.summary)

# Quick look at R^2 growth as we add variables
reg.summary$rsq
```

::: callout-tip
`regsubsets()` searches the model space and stores the *best* model for each size (1..nvmax). Use `summary()` to see a **which** matrix and metrics across sizes.
:::

### Visual diagnostics for choosing model size

```{r}
par(mfrow = c(2,2))

# RSS
plot(reg.summary$rss, xlab = "Number of Variables", ylab = "RSS", type = "l")

# Adjusted R^2 (highlight maximum)
plot(reg.summary$adjr2, xlab = "Number of Variables", ylab = "Adjusted R^2", type = "l")
best.adjr2 <- which.max(reg.summary$adjr2)
points(best.adjr2, reg.summary$adjr2[best.adjr2], col = "red", cex = 2, pch = 20)

# Cp (highlight minimum)
plot(reg.summary$cp, xlab = "Number of Variables", ylab = "Cp", type = "l")
best.cp <- which.min(reg.summary$cp)
points(best.cp, reg.summary$cp[best.cp], col = "red", cex = 2, pch = 20)

# BIC (highlight minimum)
plot(reg.summary$bic, xlab = "Number of Variables", ylab = "BIC", type = "l")
best.bic <- which.min(reg.summary$bic)
points(best.bic, reg.summary$bic[best.bic], col = "red", cex = 2, pch = 20)
```

::: callout-note
**Rule of thumb:** prefer the model size where **Adjusted R²** peaks, or **Cp/BIC** are minimized. These are proxies for **test error**.
:::

### Inspect variables in the chosen model(s)

```{r}
# Example: variables in the BIC-optimal model
coef(regfit.full, best.bic)

# regsubsets has its own plot helper:
plot(regfit.full, scale = "bic")
plot(regfit.full, scale = "adjr2")
plot(regfit.full, scale = "Cp")
plot(regfit.full, scale = "r2")
```

---

## Lab 3 — Forward Stepwise Selection

### 🎯 Learning outcomes
- Run **forward stepwise** selection using `method = "forward"`.
- Compare chosen variables to **best subset** at the same size.

```{r}
regfit.fwd <- regsubsets(Salary ~ ., data = Hitters, nvmax = 19, method = "forward")
summary(regfit.fwd)

# Compare a specific size (e.g., 7 variables) across methods
coef(regfit.full, 7)  # best subset
coef(regfit.fwd, 7)   # forward stepwise
```

::: callout-warning
Forward stepwise is **greedy**: after a variable enters, it **stays**. It evaluates far fewer models than best subset—great for speed—but it may miss the global optimum.
:::

---

## Lab 4 — Backward Stepwise Selection

### 🎯 Learning outcomes
- Run **backward stepwise** selection using `method = "backward"`.
- Understand when backward is **not** applicable.

```{r}
regfit.bwd <- regsubsets(Salary ~ ., data = Hitters, nvmax = 19, method = "backward")
summary(regfit.bwd)

# Compare at size 7 again
coef(regfit.bwd, 7)   # backward stepwise
```

::: callout-important
Backward stepwise starts from the **full model** and removes variables. It requires **n > p** so that the full least squares fit exists.
:::

---

## Lab 5 — Reading the Plots & Coefficients Like a Pro

### 🎯 Learning outcomes
- Interpret **selection plots** (BIC, Cp, Adjusted R² panels).
- Extract and report the **final model** clearly.

```{r}
# Choose a final size by any one of the criteria
k_final <- best.bic     # example: BIC-minimizer

# Coefficients and variables
final_coef <- coef(regfit.full, k_final)
final_coef

# A compact report
data.frame(term = names(final_coef), estimate = as.numeric(final_coef))
```

::: callout-tip
**Reporting tip:** Always state (i) the **criterion** used (e.g., BIC), (ii) the **model size**, and (iii) the **selected variables** with their **coefficients**. Avoid training **R²** alone—prefer Cp/BIC/Adjusted R².
:::

---

### ✅ What you should now be able to do
- Prepare data and run **best subset** and **stepwise** selection.
- Use **Adjusted R²**, **Cp**, and **BIC** to choose model size.
- Extract and communicate the **chosen variables** and **coefficients**.

> Next session we’ll compare these selections using a **validation set and cross‑validation**, and then move to **ridge** and **lasso**.
